Multi-target detection with application to cryo-electron microscopy

### 2019-02-26 inverse_problems/inverse_problems.tex

In this section, we present three numerical experiments: \TODO{menu}.

\subsection{Experiment 1}

\begin{figure}[t]
	\centering
	\includegraphics[width=1\linewidth]{heterogeneous_progressive_n12300000000_466300}
	\caption{ds}
	\label{fig:1D_example}
\end{figure}

For the 1-D experiment depicted in Figure~\ref{fig:1Dheterosignals}, we fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
is about 1/9.

This is enough noise to make cross-correlations of $y$ even with the true signals display peaks at essentially random locations, uninformative of the actual locations of the signal occurrences. Thus, we contend that it would be difficult for any algorithm to locate the signal occurrences, let alone to classify them according to which signal appears where.

Given the observation $y$, we proceed to compute the autocorrelations. The first-order autocorrelation is straightforward. For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---there are $\frac{(L-1)(L-2)}{2}$ remaining entries. We have
\begin{align*}
1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
coefficients in total. Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
\begin{align*}
K(L+1) \leq \frac{1}{2} L (L-1) + 1.
\end{align*}
Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$ (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}). \TODO{The last paragraph can be removed}
In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
\begin{multline}
\min_{\substack{\hat x_1, \ldots, \hat x_K \in \R^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2,
\label{eq:optim1D}
\end{multline}
where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).

Setting $W = L$ (as is a priori desired) is problematic because the above optimization problems appears to have numerous poor local optimizers.
Thus, we first run the optimization with $W = 2L-1$. This problem appears to have few poor local optima, perhaps because the additional degrees of freedom allow for more escape directions. Since we hope the signals estimated this way correspond to the true signals zero-padded to length $W$, we extract from each one a subsignal of length $L$ that has largest $\ell_2$-norm. This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this procedure is reliable for a wide range of experimental parameters. To solve~\eqref{eq:optim1D}, we run the trust-region method implemented in Manopt~\cite{manopt}, which allows to treat the positivity constraints on coefficients $\hat \gamma_k$. Notice that the cost function is a polynomial in the variables, so that it is straightforward to compute it and its derivatives.

>>>>

In this section, we present three numerical experiments. In the first two experiments, we consider the heterogeneous model~\eqref{eq:model}. First, with a fixed noise level, we show how the estimation quality improves as the length of the observation grows. Second, we explore how many signals can be recovered as a function of their length, in an infinite data regime where effects of the noise have been averaged out. In the last experiment, we extend our model to 2D signals.

\subsection{Experiment 1} \label{sec:XP1}

\begin{figure}[t]
	\centering
	\includegraphics[width=1\linewidth]{heterogeneous_progressive_n12300000000_466300}
	\caption{Experiment described in Section~\ref{sec:XP1}. For a fixed noise level $\sigma = 3$ and a fixed set of $K = 3$ signals of length $L = 21$, an observation $y$ of length $N = 12.3 \cdot 10^9$ is generated according to~\eqref{eq:model} in the well-separated model, with fixed density and occurrence probabilities. Each rows corresponds to one of the signals. The last column shows evolution of the relative root mean squared error in estimating each signal, as a longer and longer subset of $y$ is observed. Red dots mark the three snapshots that are illustrated in columns 1--3: red signals are the ground truth and blue signals are the estimators.}
	\label{fig:1Dheterosignals}
\end{figure}

For the experiment depicted in Figure~\ref{fig:1Dheterosignals}, we fix $K = 3$ signals of length $L = 21$: see the three red signals in the first column. Following~\eqref{eq:model}, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals (well separated). This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $\pi = (1/2, 1/3, 1/6)$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting signal-to-noise ratio of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
is about 1/9.

Visually, the noise dominates the signal to the point that it is challenging to detect occurrences. More precisely, the cross-correlations of $y$ even with the true signals presents peaks at essentially random locations, apparently uninformative of the actual locations of the signal occurrences. Thus, we contend that it would be difficult for any algorithm to locate the signal occurrences, let alone to classify them according to which signal appears where.

%Given the observation $y$, we proceed to compute the autocorrelations. The first-order autocorrelation is straightforward. For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---there are $\frac{(L-1)(L-2)}{2}$ remaining entries. We have
%\begin{align*}
%1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
%\end{align*}
%coefficients in total. Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
%\begin{align*}
%K(L+1) \leq \frac{1}{2} L (L-1) + 1.
%\end{align*}
%Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$ (compare with~\cite{boumal2017heterogeneous,bandeira2017estimation}).

Our aim is to investigate how accurately we can estimate the signals as a function of the observation length. To this end, we consider a growing part of the observation $y$. For each length, we compute autocorrelations on that part, then we go on to estimate the signals from these ``mixed'' quantities. In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

Having computed the autocorrelations of interest, we estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
\begin{multline}
	\min_{\substack{\hat x_1, \ldots, \hat x_K \in \R^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2,
	\label{eq:optim1D}
\end{multline}
where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).

Setting $W = L$ (as is a priori desired) is problematic because the above optimization problems appears to have numerous poor local optimizers. Thus, we first run the optimization with $W = 2L-1$. This problem appears to have few poor local optima, perhaps because the additional degrees of freedom allow for more escape directions. Since we hope the signals estimated this way correspond to the true signals zero-padded on either side to length $W$, we extract from each one a subsignal of length $L$ that has largest $\ell_2$-norm. This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this procedure is reliable for a wide range of experimental parameters. To solve~\eqref{eq:optim1D}, we run the trust-region method implemented in Manopt~\cite{manopt}, which allows to treat the positivity constraints on coefficients $\hat \gamma_k$. Notice that the cost function is a polynomial in the variables, so that it is straightforward to compute it and its derivatives.

In Figure~\ref{fig:1Dheterosignals}, we find that the signals can be recovered with good accuracy despite the noise levels which seemingly preempts location and classification. We also note that the amount of data required to produce these good estimations is large. Furthermore, as illustrated here and as we have observed in numerous experiments, signals with more variations (such as the third signal in this experiment which was generated once from a Gaussian distribution) are easier to estimate accurately than more regular signals (in this case, despite the fact that the third signal occurs less frequently than the others).
